Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-07 Paper/paper.tex
para to transitioning from nsbh general to paper

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1''
mid-September 2015, operating at a factor of $3-4$ higher gravitational-wave (GW) 
strain sensitivity than their first-generation counterparts~\cite{Shoemaker2009}.
% 
During O1, the first terrestrial detection of gravitational waves was made, as
was announced earlier in 2016~\cite{LIGOVirgo2016a}. Emitted more than a billion
year ago, these waves herald the era of observational gravitational wave (GW)
astronomy as they pass through Earth.
% 
Over a $3$-month O1, aLIGO had access to more binary mergers than initial-LIGO
over its entire coincident operation period. Both of aLIGO instruments will be
further upgraded in phases to reach their design sensitivity by $2018-19$, at 
which point we may expect to observe $\sim 70$ binary mergers a 
year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or neutron 
stars (NS), remain the most interestings targets for GW based observation with
aLIGO. NSBH binaries form an important class of compact binaries, and we expect
to see $\mathcal{O}(10)$ of these every year with design-sensitivity 
aLIGO~\cite{Abadie:2010cf}.
% 
NSBH mergers are of interest for various reasons. For instance, they have 
been associated with (as the progenitors of) short Gamma-ray Bursts
(GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Plausible mechanisms have been proposed for such a 
scenario~\cite{Paschalidis2014,Ciolfi:2014yla}. A joint detection of GWs
from an NSBH system with a GRB would validate this hypothesis.
% 
Another question that NSBH mergers can be used to probe is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron stars. 
% 
The gravitational-waves emitted by NSBH binaries carry subtle hints of the 
matter structure of the neutron star. During early inspiral, the (weak) tidal
field of the companion hole deforms the star, thereby exciting its resonant
oscillation modes. This drains energy from the orbit resulting in an
increase of the inspiral rate. We know the leading and next-to-leading
order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN. Closer to merger, 
the neutron star could get disrupted by the tidal field of its companion.
Once disrupted, the neutron star material may form an accretion disk around 
the black hole. Most of this matter disk spirals in within a few 
milliseconds, with a small fraction becoming unbounded and gets 
ejected~\cite{Foucart2015,Lovelace:2013vma,Deaton2013,Foucart2012,
Paschalidis2014} (under favorable conditions, and possible accompanied by 
gamma-ray emission).
The emitted GW signal is, as a result, heavily damped post-disruption, 
and the quasi-normal modes of the resulting black hole are not excited. Such a
GW signal will be qualitatively very different from that from an equivalent
binary black hole (BBH) merger. The primary goal of this work is to understand
what it requires to optimally extract information about neutron star structure
from such NSBH signals.

However, studies have shown that for mass-ratio $\mbh/\mns\gtrsim 5$ the BH
will need to have fairly large and positively-aligned (with the orbit) spins
in order for its tidal field to be strong enough to disrupt the NS before the
innermost stable circular orbit (ISCO) is reached~\cite{Foucart:2013psa}.
% 
It is worth noting
that not all point-particle terms in PN theory are known up to such high order,
and therefore we need to be careful when using PN waveforms to measure tidal
effects.
%

Conventional astronomical methods have, to date, made observations of ~2500
NSs with masses between $1.25-2.1 M_\odot$~\cite{Lyne:2004cj,
Demorest:2010bx,2013Sci...340..448A,atnfcatalog,mcgillmagnetarcatalog,
stellarcollapsemass}, although
tightly clustered within $1.35\pm0.5M_\odot$~\cite{stellarcollapsemass}.
%
Observations of over 2500 NSs have shown that the spin of NSs
$|\vec{\chi}_\mathrm{NS}|$ have magnitudes that are $<1\%$~\cite{Miller:2014aaa}.
%
On the other hand, indirect observations of stellar-mass BHs place their
masses between $5-35M_\odot$, with their spin angular momenta 
$|\vec{\chi}_\mathrm{BH}|$ ranging from small to nearly extremal (Kerr) values
(see, e.g., Refs.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for 
examples of nearly extremal estimates of BH spins, Refs.~\cite{McClintock:2013vwa,
Reynolds:2013qqa} for recent reviews of astrophysical BH spin measurements,
and Figure 5 of Ref.~\cite{Miller:2014aaa} for a comparison of NS and BH spins).

The difficulty in discerning the matter effects in an NSBH merger GW signal
lies in distinguishing it reliably from a BHBH merger signal. The earlier the
NS disrupts, the more orbits there are over which the effect of NS disruption
is visible. Where exactly does disruption happen is governed both by the 
mass-ratio of the binary, as well as the spin on the BH. Larger BHs (with
$q=m_\mathrm{BH}/m_\mathrm{NS}\gtrsim 7$), as well as BHs with large retrograde
spins ($\chi_\mathrm{BH} \lesssim 0$), tend to swallow the NS whole due to a
rapid merger (i.e. without it disrupting before the inner-most stable circular
orbit (ISCO) is reached). On the other hand, binaries with mass-ratios 
$q=2-5$ with large pro-grade $\chi_\mathrm{BH}$ can disrupt much before 
merger, leaving their imprint in the GW signal emitted close to $1.7-2$~kHz.
%

@@

@@

In Ref.~\cite{Lackey:2013axa}, the authors presented a suite of $134$ numerical
relativity simulations of neutron stars inspiraling into spinning black holes. Amongst
the varied quantities were (a) the equation of state for the NS, (b) NS mass, 
(c) mass-ratio and (d) black hole spins, taken as aligned with the orbit.
A total of $21$ two-parameter EoSs were used for these simulations, which also 
had $q\in[2, 5]$, $\chibh\in[-0.5, +0.75]$, and 
$m_\mathrm{NS}\in[1.20, 1.45]M_\odot$. Based on this suite of simulations, the
authors calibrated a frequency-domain waveform model for the inspiral and merger phasing of
NSBH coalescences. The GW strain predicted by this model (henceforth ``LEA''),
can be written as
% 
\begin{equation}
 \tilde{h}_\mathrm{NSBH}(f, \vec{\theta}) = \tilde{h}_\mathrm{BBH}(f, \vec{\theta} - \{\lambdans\})\,A(f, \vec{\theta})\,e^{\ii \Delta\Phi(f, \vec{\theta})},
\end{equation}
where $\vec{\theta}:=\{\mbh, \mns, \chibh,\chins=0, \lambdans\}$ are the intrinsic binary
parameters. The first term on the right hand side is an underlying
BBH waveform model that is used as a base. In the original construction of LEA,
this was taken to be the SEOBNRv1 model~\cite{Taracchini:2012} of the 
Effective-one-body (EOB) family~\cite{Buonanno99}. The second term changes the 
amplitude of the BBH model
to match that of an NSBH merger of otherwise identical parameters, with NS-matter
effects parametrized by $\lambdans$. During inspiral this term is set to
unity, but is a sensitive function of $\lambdans$ close to merger. The third 
term does the same as the second, but to the waveform phasing. During inspiral,
$\Delta\Phi(f, \vec{\theta})$ is set to the PN tidal phasing corrections,
at the leading and next-to-leading orders~\cite{Vines2011}. Close to merger,
additional phenomenological terms are added and calibrated to the $134$ available
NR simulations.

@@

The joint inferred probability distribution of all parameters $\vec{\theta}\cup\{\lambdans\}$ 
is given by
\begin{equation}
 p(\vec{\theta}, \lambdans | d_n, K) = \dfrac{p(d_n|\vec{\theta}, \lambdans, K)\,p(\vec{\theta}, \lambdans | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\theta}, \lambdans | K)$ is the prior probability for binary parameters of taking 
particular values. In this paper, we use a uniform prior on the individual masses within 
pre-chosen ranges, as well as a uniform prior on the spin of the black hole. We set neutron star's
spin identically $=0$, and its tidal deformability parameter $\lambdans$ to be equally likely
between $[0, 4000]$. The prior probability of obtaining a particular realization of data
$p(d_n|K)$ is absorbed into the overall normalization. Finally, the first term in the numerator
$p(d_n|\vec{\theta}, \lambdans, K)$ is the likelihood of obtaining the given stretch of 
data $d_n$ if we assume that a signal parameterized by $\vec{\theta}\cup\{\lambdans\}$ is buried
in it, i.e.
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\theta}, \lambdans) \equiv p(\vec{\theta}, \lambdans | K) = \mathcal{N} \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\theta}, \lambdans)$ is our waveform model, and the detector noise
weighted inner-product $\langle a|b\rangle$ is defined as 
$$
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
$$
where $\tilde{a}$ means the Fourier transform of the finite time series $a(t)$, and 
$S_n(|f|)$ is the one-sided amplitude spectrum of detector noise. The definition of $\LL$ in 
Eq.~\ref{eq:likelihood} assumes that the instrument noise is colored Gaussian.

>>>>

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1''
mid-September 2015, operating at a factor of $3-4$ higher gravitational-wave
(GW) strain sensitivity than their first-generation 
counterparts~\cite{Shoemaker2009}.
% 
During O1, the first terrestrial detection of gravitational waves was made, as
was announced earlier in 2016~\cite{LIGOVirgo2016a}. Emitted more than a
billion year ago, these waves herald the era of observational gravitational
wave (GW) astronomy as they pass through Earth.
% 
Over the $3$-months of O1, aLIGO had access to more binary mergers than 
initial-LIGO over its entire coincident operation period. Both of aLIGO 
instruments will be further upgraded in phases to reach their design 
sensitivity by $2018-19$, at which point we may expect to observe $\sim 70$ 
binary mergers a year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or neutron 
stars (NS), remain the most interestings targets for GW based observation with
aLIGO. NSBH binaries form an important sub-class of compact binaries, and we
expect to observe $\mathcal{O}(10)$ of these every year with design 
sensitivity aLIGO~\cite{Abadie:2010cf}.
% 
NSBH mergers are of interest for various reasons. For instance, they have 
been associated with (as the progenitors of) short Gamma-ray Bursts
(GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, close to merger the neutron star
can get disrupted by the tidal field of its companion hole. Once disrupted,
the neutron star material may form an accretion disk around the black hole. 
Most of the disk matter spirals in within a few milliseconds, with a small 
fraction becoming unbounded and getting ejected, with a possibled simultaneous
burst of gamma-ray emission)~\cite{Foucart2015,Lovelace:2013vma,Deaton2013,
Foucart2012,Shibata:2005mz,Paschalidis2014}. A coincident detection of GWs
from an NSBH system with a GRB would therefore provide us insights into the
central engine powering GRBs. 
% 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron stars. This is what we will focus on in this paper.
% 
The gravitational-waves emitted by NSBH binaries carry subtle hints of the 
matter structure of the neutron star. During early inspiral, the (weak) tidal
field of the companion hole deforms the star, thereby exciting its resonant
oscillation modes. This drains energy from the orbit resulting in an
increase of the inspiral rate. We know the leading and next-to-leading
order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. Close to
merger, as the NS gets disrupted and forms long tails that feed into the 
companion black hole, the quadrupole moment of the system falls monotonically.
As a result, the emitted GWs are subdued significantly post-disruption
(as compared to what they would be for a BBH),
and the quasi-normal modes of the resulting black hole are not 
excited~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart2015,
Pannarale:2015jia}.
% 
% 
The difficulty in discerning NS matter effects in an NSBH coalescence from its
GW signal lies in distinguishing it reliably from a BBH signal, since the 
difference between the two is either subtle (during inspiral) or short-lived
(disruption and merger). The earlier in inspiral the NS disrupts, the more
orbits there are over which its effect is visible. The mechanism of disruption
is governed by the nature of the BH, i.e. its mass and spin. More massive BHs 
(with $m_\mathrm{BH}\gtrsim 7\, m_\mathrm{NS}$), as well as BHs with
(large) retrograde spins $(\chi_\mathrm{BH}\in[-1. 0])$, tend to swallow the 
NS whole due to a rapid merger (i.e. without it disrupting before the
inner-most stable circular orbit (ISCO) is reached). The quasi-normal
oscillation modes of the final black hole's horizon are also 
excited~\cite{Foucart:2013psa}. Therefore, the only available information 
comes from the inspiral rate modification. On the other hand, less massive
black holes spinning in the prograde sense (with mass-ratio 
$q\in[2, 5]$, and $\chi_\mathrm{BH}\geq 0$) can cause their companion star to 
disrupt much before merger, leaving a strong imprint on the emitted GW 
signal~\cite{Shibata:2007zm,2010PhRvD..81f4026F,Lovelace:2013vma,
Kawaguchi:2015}.

A large fraction of past work aimed at measuring NS matter effects from GW
signals has consisted of inquiries about binary neutron stars (BNS)~\cite{
Lee1999a,Lee1999b,Lee2000,oechslin:07,Read:2008iy,Markakis:2010mp,Markakis:2011vd,
stergioulas:11,East:2011xa,Lackey2014,Wade:2014vqa,Bauswein:2014qla}. While single 
observations are not as informative, Bayesian studies have shown that with
$20-30$ BNS observations, aLIGO would begin to distinguish between hard,
moderate and soft candidate equations of state~\cite{DelPozzo:13,
Chatziioannou:2015uea,Agathos:2015a}. These studies rely on the PN 
description of BNS dynamics, which is reasonable since BNSs merge at very high
frequencies and most of their signal power resides in inspiral.
% 
On the other hand, NSBHs merge at relatively lower frequencies than BNS, 
and their disruption during late-inspiral occurs at frequencies where aLIGO
has relatively more resolving power. This, combined with a shorter inspiral,
means that the merger portion of NSBH signals is relatively more valuable for
measuring $\lambdans$. Some past NSBH studies aimed at measuring tidal effects,
however, have used PN inspiral-only waveforms~\cite{Maselli:2013rza}. In doing
so, however, (i) they ignore the merger signal which could contain significant
information for NSBHs, and (ii) the errors due to unknown vaccum terms in PN 
waveforms could easily dominate over the tidal terms 
themselves~\cite{Yagi:2014}, since there are unknown point-particle terms at 
orders {\it lower} than that at which tidal effects enter. Other studies that
use complete numerical simulations or derived models either use Fisher matrix 
to estimate $\lambdans$ measurement error~\cite{Lackey2011,Lackey:2013axa},
which is not reliable at realistic signal-to-noise ratios 
(SNR)~\cite{Vallisneri:2007ev}, or are limited in the binary parameter space
they sample~\cite{Foucart:2013psa}.

@@

As this line of inquiry relies mainly on 
classifying a GW signal as an NSBH binary, it can be pursued using non-tidal
waveform models~\cite{}.

% Conventional astronomical methods have, to date, made observations of ~2500
% NSs with masses between $1.25-2.1 M_\odot$~\cite{Lyne:2004cj,
% Demorest:2010bx,2013Sci...340..448A,atnfcatalog,mcgillmagnetarcatalog,
% stellarcollapsemass}, although
% tightly clustered within $1.35\pm0.5M_\odot$~\cite{stellarcollapsemass}.
% %
% Observations of over 2500 NSs have shown that the spin of NSs
% $|\vec{\chi}_\mathrm{NS}|$ have magnitudes that are $<1\%$~\cite{Miller:2014aaa}.
% %
% On the other hand, indirect observations of stellar-mass BHs place their
% masses between $5-35M_\odot$, with their spin angular momenta 
% $|\vec{\chi}_\mathrm{BH}|$ ranging from small to nearly extremal (Kerr) values
% (see, e.g., Refs.~\cite{McClintockEtAl:2006,Miller:2009cw,Gou:2014una} for 
% examples of nearly extremal estimates of BH spins, Refs.~\cite{McClintock:2013vwa,
% Reynolds:2013qqa} for recent reviews of astrophysical BH spin measurements,
% and Figure 5 of Ref.~\cite{Miller:2014aaa} for a comparison of NS and BH spins).
%

@@

In Ref.~\cite{Lackey:2013axa}, the authors presented a suite of $134$ numerical
relativity simulations of neutron stars inspiraling into spinning black holes. Amongst
the varied quantities were (a) the equation of state for the NS, (b) NS mass, 
(c) mass-ratio and (d) black hole spins, taken as aligned with the orbit.
A total of $21$ two-parameter EoSs were used for these simulations, which also 
had $q\in[2, 5]$, $\chibh\in[-0.5, +0.75]$, and 
$m_\mathrm{NS}\in[1.20, 1.45]M_\odot$. Based on this suite of simulations, the
authors calibrated a frequency-domain waveform model for the inspiral and merger phasing of
NSBH coalescences. The GW strain predicted by this model (henceforth ``LEA''),
can be written as
% 
\begin{equation}
 \tilde{h}_\mathrm{NSBH}(f, \vec{\theta}, \lambdans) = \tilde{h}_\mathrm{BBH}(f, \vec{\theta})\,A(f, \vec{\theta}, \lambdans)\,e^{\ii \Delta\Phi(f, \vec{\theta}, \lambdans)},
\end{equation}
where $\vec{\theta}:=\{\mbh, \mns, \chibh,\chins=0\}$ are the intrinsic binary
parameters. The first term on the right hand side is an underlying
BBH waveform model that is used as a base. In the original construction of LEA,
this was taken to be the SEOBNRv1 model~\cite{Taracchini:2012} of the 
Effective-one-body (EOB) family~\cite{Buonanno99}. The second term changes the 
amplitude of the BBH model
to match that of an NSBH merger of otherwise identical parameters, with NS-matter
effects parametrized by $\lambdans$. During inspiral this term is set to
unity, but is a sensitive function of $\lambdans$ close to merger. The third 
term does the same as the second, but to the waveform phasing. During inspiral,
$\Delta\Phi(f, \vec{\theta}, \lambdans)$ is set to the PN tidal phasing corrections,
at the leading and next-to-leading orders~\cite{Vines2011}. Close to merger,
additional phenomenological terms are added and calibrated to the $134$ available
NR simulations.

@@

The joint inferred probability distribution of all parameters 
$\vec{\Theta}:=\vec{\theta}\cup\{\lambdans\}$ 
is given by
\begin{equation}
 p(\vec{\Theta} | d_n, K) = \dfrac{p(d_n|\vec{\Theta}, K)\,p(\vec{\Theta} | K)}{p(d_n|K)}.
\end{equation}
Here, $p(\vec{\Theta} | K)$ is the prior probability for binary parameters of taking 
particular values. In this paper, we use a uniform prior on the individual masses within 
pre-chosen ranges, as well as a uniform prior on the spin of the black hole. We set neutron star's
spin identically $=0$, and its tidal deformability parameter $\lambdans$ to be equally likely
between $[0, 4000]$. The prior probability of obtaining a particular realization of data
$p(d_n|K)$ is absorbed into the overall normalization. Finally, the first term in the numerator
$p(d_n|\vec{\theta}, \lambdans, K)$ is the likelihood of obtaining the given stretch of 
data $d_n$ if we assume that a signal parameterized by $\vec{\theta}\cup\{\lambdans\}$ is buried
in it, i.e.
\begin{equation}\label{eq:likelihood}
 \LL(\vec{\Theta}) \equiv p(\vec{\Theta} | K) = \mathcal{N} \mathrm{exp}[- \langle d_n - h | d_n - h\rangle ],
\end{equation}
where $h\equiv h(\vec{\Theta})$ is given by our waveform model LEA+, and the
detector noise weighted inner-product $\langle a|b\rangle$ is defined as 
$$
\langle a|b\rangle \equiv 4\,\mathrm{Re}\left[\int_0^\infty \dfrac{\tilde{a}(f) \tilde(b)(f)^*}{S_n(|f|)}\,\D f\right],
$$
where $\tilde{a}$ means the Fourier transform of the finite time series $a(t)$, and 
$S_n(|f|)$ is the one-sided amplitude spectrum of detector noise. The definition of $\LL$ in 
Eq.~\ref{eq:likelihood} assumes that the instrument noise is colored Gaussian.
